---
title: "Beautiful Graphics"
subtitle: "Notes to the book by Alboukadel Kassambara"
author:
- name: "Georg Maubach"
  affiliation: "Veikko Consulting"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    self_contained: yes
    toc: no
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    toc_title: "Table of Contents"
  pdf_document:
    keep_md: yes
    number_sections: yes
    toc: no
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_title: "Table of Contents"
  word_document:
    toc: no
    toc_depth: 3
keywords: graphics, notes, ggplot2
linestretch: 1
fontsize: 11pt
tags:
- graphics
- notes
- ggplot2
abstract: |
  This document contains notes while reading the book.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The is http://www.sthda.com/english/wiki/ggplot2-essentials.

# Introduction
## Installing and Loading ggplot2
```{r eval = FALSE}
warning("Use type = 'source' only if you have configured a C++ compiler.")

local({
  r = getOption("repos")
  r["CRAN"] = "https://cran.rstudio.com/"
  r["CRANextra"] <- "http://www.stats.ox.ac.uk/pub/RWin"
  r["CRAN_de"] <- "https://cran.uni-muenster.de/"
  options(repos = r)
})

install.packages(
  pkgs = "ggplot2", 
  dependencies = TRUE,
  type = "source")

library(ggplot2)
```

- - -

# Introduction to ggplot2
## Terms
A plot can be divided into **3** fundumental parts:
**Plot = data + Aesthetics + Geometry**.

- **data**: a data frame
- **Aesthetics**: describe the mapping of variables to visual properties of geometry.
    1. indicate the x and y variables,  
    2. control color, size, shape of points, etc. 
- **Geometry**: describe the type of graphic, e.g. histogram, box plot, line plot, scatter plot, etc.

Geometry is defined in geom_*(). Geoms are called layers cause the can occur multiple times and put on top of each other.

### Further reading
- http://docs.ggplot2.org/0.9.3/aes.html
- http://docs.ggplot2.org/current/vignettes/ggplot2-specs.html

## Data format and preparation
- **Data must be a data frame, containing all information to make a ggplot graphic.**  
- **Data should be tidy, i. e. columns should be variables, rows should be observations.**  

```{r}
# mtcars
## Load data
data(mtcars)
df <- mtcars[ , c("mpg", "cyl", "wt")]

## Convert dyl to a factor variable
df$cyl <- as.factor(df$cyl)

## Print a sample of data
head(mtcars)

#----------------------------------------------------------
library(dplyr)

set.seed(1234)
wdata <- data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))

head(wdata)

mu <- wdata %>%
  group_by(sex) %>%
  summarise(grp.mean = mean(weight))

head(mu)
```

## ggplot Basics
### ggplot basic example
```{r}
library(ggplot2)
# Basic scatter plot
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# Change the font size and shape
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
   geom_point(size = 1.5, shape = 18)
```

### aes_string()
aes_string() generates aesthetics from a string. This is particularly useful when writing functions that create plots cause strings can be used to define aesthetic mappings. Otherwise it would be needed to use a substitute to generate a call to aes().

```{r}
ggplot(data = mtcars, aes_string(x = "wt", y = "mpg")) +
   geom_point(size = 1.5, shape = 18)

ggpoints <- function(data, x_name, y_name)
{
  p <- ggplot(data = data, aes_string(x_name, y_name)) +
    geom_point(color = "red") +
    geom_smooth()
  
  return(p)
}

ggpoints(data = mtcars, x_name = "wt", y_name = "mpg")
```

### Visualise transformations of original dataset
Some plots visualise a transformation on the original data set. In this case, an alternative way to build a layer is to use stat_*() functions.

```{r}
set.seed(1234)
wdata = data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))

print("Use geometry function")
ggplot(wdata, aes(x = weight)) + geom_density()

print("Use stat function")
ggplot(wdata, aes(x = weight)) + stat_density()
```

### Layers using geoms
Each plot consists of one or more layers. Layers are defined using geom_*().

#### Using the same data and same aesthetic mapping in one plot
```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point() +  # to draw points
  geom_line()     # to draw a line
```

#### Using different data and mappings for different layers
```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point() +  # to draw points from whole data set
  geom_line(data = head(mtcars), color = "red")  # to draw a line from a subset of data
```

### Calculations in aes()
```{r}
ggplot(data = mtcars, aes(x = log2(wt), y = log2(mpg))) +
  geom_point()
```

### Saving plots
- Plots can be stored as a variable and printed using print().  
- last_plot(): returns last plot modified  
- ggsave("plot.png", width = 5, height = 5): save the last plot to the current working directory  

#### Saving directly from screen
```{r}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
ggsave("my_plot1.pdf")  # to PDF
ggsave("my_plot1.png")  # to PNG
```

#### Saving graphic objects
```{r}
pdf("my_plot2.pdf")
my_plot <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
print(my_plot)
dev.off()

png("my_plot2.png")
my_plot <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
print(my_plot)
dev.off()
```
- - -

# Plotting **ONE** variable (discrete or continuous)

```{r}
library(dplyr)

set.seed(1234)
wdata <- data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))

head(wdata)

mu <- wdata %>%
  group_by(sex) %>%
  summarise(grp.mean = mean(weight))

head(mu)
```

```{r}
my_plot <- ggplot(wdata, aes(x = weight))
```

## Overview of possible graphics

### Discrete Variables

- geom_bar():       bar plot

### Continuous Variables

- geom_area():      area plot
- geom_density():   density plot
- geom_dotplot():   dot plot
- geom_freqpoly():  frequency polygon
- geom_histogram(): histogram plot
- geom_ecdf():      empirical cumulative density function
- stat_qq():        quantile-quantile plot

## Bar plot
### Basics
- Definition: Bar plot can be used to visualise one discrete variable. A count of each level is plotted.
- Key functions: geom_bar()
- Alternvative function: stat_count()
- Key arguments: alpha, color, fill, linetype, size

### Example
```{r}
data(mpg)
ggplot(mpg, aes(fl)) +
  geom_bar(fill = "steelblue") +
  theme_minimal()
```


## Area Plot
### Basics
- Definition: Area plots are the continuous analog to a stacked bar chart.
- Key function: geom_area()
- Alternative function: stat_bin()
- Key arguments: alpha, color, fill, linetype, size

### Example 1: y Values Corresponding to the Count of x Values

```{r}
my_plot + geom_area(stat = "bin", color = "black", fill = '#00AFBB')
```

### Example 2: y Values Corresponding to the Density of x Values
```{r}
my_plot + geom_area(aes(y = ..density..), stat = "bin")
```

### Comparison of Bar Plot and Area Plot
```{r}
data("diamonds")
head(diamonds)

p <- ggplot(diamonds, aes(x = price, fill = cut))

p + geom_bar(stat = "bin")

p + geom_area(stat = "bin")
```

## Density Plot
### Basics
- Density plots are useful to visualise the distribution of a continuous variable.
- Key function: geom_density()
- Alternative function: stat_density()
- Key arguments: alpha, color, fill, linetype, size
- References
    1. https://en.wikipedia.org/wiki/Density_estimation
    2. https://de.wikipedia.org/wiki/Kerndichtesch%C3%A4tzer

### Example 1
```{r}
my_plot + geom_density()

my_plot + geom_density(color = "black", fill = "grey") +
  geom_vline(aes(xintercept = mean(weight)),
             color = "#FC4E07", 
             linetype = "dashed",
             size = 1)
```

### Groupings
#### Example 1: Line Color by Sex
```{r}
my_plot + geom_density(aes(color = sex))
```

#### Example 2: Fill Color by Sex, semi-transparent fill (alpha = 0.4)
```{r}
my_plot + geom_density(aes(fill = sex), 
                       alpha = 0.4)
```

#### Example 3: Line Color and Mean Line by Sex
```{r}
head(mu)
my_plot + geom_density(aes(color = sex), 
                        alpha = 0.4) +
  geom_vline(data = mu,
             aes(xintercept = grp.mean, 
                 color = sex),
             linetype = "dashed")
```

### Graphic Property Control
#### Basics
- scale_color_manual(), scale_fill_manual(): use custom colors
- scale_color_brewer(), scale_fill_brewer(): use color palette from RColorBrewer package
- scale_color_grey(), scale_fill_grey():     use grey color palettes

#### Manually Defined Lines
##### Example 0: Create Basic Graphic
```{r}
my_plot_2 <- my_plot + 
  geom_density(aes(color = sex)) +
  geom_vline(data = mu,
             aes(xintercept = grp.mean, 
                 color = sex),
             linetype = "dashed") +
  theme_minimal()
```

##### Example 1: Manually Defined Color
```{r}
my_plot_2 + scale_color_manual(values = c("#999999", "#E69F00"))
```

##### Example 2: Colors Using RColorBrewer Palettes
```{r}
my_plot_2 + scale_color_brewer(palette = "Paired")
```

##### Example 3: Grey Scale
```{r}
my_plot_2 + scale_color_grey()
```

#### Manually Defined Fills
##### Example 0: Create Basic Graphic
```{r}
my_plot_3 <- my_plot + 
  geom_density(aes(fill = sex), alpha = 0.4) +
  theme_minimal()
```

##### Example 1: Manuall Fill
```{r}
my_plot_3 + scale_fill_manual(
  values = c("#999999", "#E69F00"))
```

##### Example 2: Fill with RColorBrewer Palettes Colors
```{r}
my_plot_3 + scale_fill_brewer(palette = "Dark2") +
  theme_minimal()
```

##### Example 3: Fill with Grey Scale Colors
```{r}
my_plot_3 + scale_fill_grey() +
  theme_minimal()
```

## Dot Plot
### Basics
- Definition: Each observation in a dot plot represents a dot. Dot plots are used to visualise one variable.
Key function: geom_dotplot()
Alternative function: none
Key arguments: alpha, color, fill, dotsize

### Example
```{r}
library(ggplot2)
set.seed(1234)
wdata <- data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))
head(wdata)

myplot <- ggplot(wdata, aes(x = weight))
myplot + geom_dotplot(
  aes(fill = sex))
```

## Frequency Polygon
### Basics
- Definition: Frequency Polygons are used to visualise the distribution of a continuous variable. It is similar to histogram plots but instead of using bars frequency polygons use lines.
Key function: geom_freqpoly()
Alternative function: stat_bin()
Key arguments: alpha, color, linetype, size

### Example 1
```{r}
library(ggplot2)
set.seed(1234)
wdata <- data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))
head(wdata)
```

#### Basic plot
```{r}
myplot <- ggplot(wdata, aes(x = weight))
myplot + geom_freqpoly(bins = 30) +
  theme_minimal()
```

#### Change color and linetype by sex, use custom color palettes
```{r}
myplot + geom_freqpoly(
  aes(
    color = sex,
    linetype = sex)) +
  scale_color_manual(values = c("#999999", "#E69F00")) +
  theme_minimal()
```

#### Plot density on y-axis
```{r}
myplot + geom_freqpoly(aes(y = ..density..))
```

## Histogram Plot
### Basics
- Definition: Histogram plots represent a distribution of a continuous variable by dividing into bins and counting the number of observations in each bin.
- Key function: geom_histogram()
- Alternative function: stat_bin()
- Key arguments: alpha, color, fill, linetype, size
- Adjustments: "identity" (or position_identity()), "stack" (or position_stack()), "dodge" (or position_dodge())
- Defaults: "stack", bins = 30, y axis = count

#### Example 1: y Values Corresponding to the Count of x Values
##### Basic plot
```{r}
my_plot + geom_histogram()
```

###### Change the number of bins
```{r}
my_plot + geom_histogram(bins = 50)
```

###### Change line color and fill color, add mean line
```{r}
my_plot + 
  geom_histogram(color = "black", fill = "grey") +
  geom_vline(
    aes(xintercept = mean(weight)),
    color = "#FC4E07",
    linetype = "dashed",
    size = 1)
```

**Note**
Instead of changing the number of bins with bins = <number> the bindwidth can be changed using binswidth = <number>.

#### Example 2: y Values Corresponding to the density of x Values
```{r}
my_plot + geom_histogram(aes(y = ..density..))
```

### Change Colors by Groups
#### Change line colors by sex
```{r}
my_plot + geom_histogram(
  aes(color = sex), 
  fill = "white",
  position = "stack")  # "stack" is default
```

#### Position adjustment "identity" (overlaid)
```{r}
my_plot + 
  geom_histogram(
    aes(color = sex),
    fill = "white",
    alpha = 0.6,
    position = "identity")
```

#### Position adjustment: "dodge" (interleaved) and mean lines by sex
```{r}
my_plot + 
  geom_histogram(
    aes(color = sex),
    fill = "white",
    position = "dodge") +
  geom_vline(
    data = mu,
    aes(xintercept = grp.mean,
        color = sex),
    linetype = "dashed")
```

### Graphic Property Control
#### Basics
- scale_color_manual(), scale_fill_manual(): use custom colors  
- scale_color_brewer(), scale_fill_brewer(): use color palette from RColorBrewer package  
- scale_color_grey(), scale_fill_grey(): use grey color palettes  

#### Manually Defined Lines
##### Example 1: Change outline color manually
```{r}
my_plot + 
  geom_histogram(
    aes(color = sex),
    fill = "white",  # fill IN geom NOT in aes !!!
    alpha = 0.4,
    position = "identity") +
  scale_color_manual(values = c("#00AFBB", "#E7B800"))
```

##### Example 2: Change fill and outline color manually
```{r}
my_plot + 
  geom_histogram(
    aes(color = sex,
        fill = sex),  # fill NOT in geom BUT in aes !!!
    alpha = 0.4,
    position = "identity") +
  scale_fill_manual(values = c("#00AFBB", "#E7B800")) +
  scale_color_manual(values = c("#00AFBB", "#E7B800"))
```

### Combine Histogram and Density Plots
- Plot histogram with density values on y-axis (instead of count values).
- Add density plot with transparent density plot.

#### Histogram with Density Plot
- Histogram with transparent filled density plot overlaid.

```{r}
my_plot + 
  geom_histogram(
    aes(y = ..density..),
    color = "black",  # color IN geom NOT in aes
    fill = "white") +  # fill in geam NOT in aes
  geom_density(
    alpha = 0.2,
    fill = "#FF6666")
```

#### Histogram with color by groups
- Histogram with line density plot overlaid.

```{r}
my_plot + 
  geom_histogram(
    aes(
      y = ..density..,
      color = sex,  # color NOT in geom BUT in aes
      fill = sex),  # fill  NOT in geom BUT in aes
    alpha = 0.5,
    position = "identity") + 
  geom_density(
    aes(color = sex),
    size = 1)
```

## ECDF plots
### Basics
- Definition: ECDF plots report for any given number the percent of individuals that are below that threshold.
- Key function: stat_ecdf()
- Alternative function: none
- Key arguments: alpha, color, linetype, size

### Examples
```{r}
library(ggplot2)
set.seed(1234)
wdata <- data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))
head(wdata)

my_plot <- ggplot(wdata, aes(x = weight))
my_plot + stat_ecdf(geom = "point")
my_plot + stat_ecdf(geom = "step")
```

## QQ Plot
### Basics
- Definition: QQ plots are used to check whether a given data follows normal distribution. 
- Key function: stat_qq()
- Alternvative function: qplot()
- Key arguments: alpha, color, shape, size

### Examples
```{r}
data(mtcars)
mtcars$cyl <- as.factor(mtcars$cyl)
head(mtcars[ , c("mpg", "cyl")])
my_plot <- ggplot(mtcars, aes(sample = mpg))

my_plot + stat_qq()
```

```{r}
my_plot + stat_qq(
  aes(
    shape = cyl, color = cyl))  + 
    scale_color_manual(
      values = c("#00AFBB", "#E7B800", "#FC4E07"))
```



