---
title: "Kassambra_Graphics_Notes"
author: "Georg Maubach"
date: "19 Februar 2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages(pkgs = "ggplot2", dependencies = TRUE, type = "source")
library(ggplot2)
```

# Chapter 2: Introduction to ggplot2

## Definition of plots

**Plot = data + Aesthetics + Geometry**

- data: a data frame  
- Aesthetics:
    1. indicate the x and y variables,  
    2. control color, size, shape of points, etc.  
- Geometry: defines the type of graphic, e. g. histogram, box plot, etc.

Geometry is defined in geom_*(). Geoms are called layers cause the can occur multiple times and put on top of each other.

## Output plots

- Plots can be stored as a variable and printed using print().  
- last_plot(): returns last plot modified  
- ggsave("plot.png", width = 5, height = 5): save the last plot to the current working directory  

## Data format and preparation

- **Data must be a data frame, containing all information to make a ggplot graphic.**  
- **Data should be tidy, i. e. columns should be variables, row should be observations.**  

```{r}
# Load data
data(mtcars)
df <- mtcars[ , c("mpg", "cyl", "wt")]

# Convert dyl to a factor variable
df$cyl <- as.factor(df$cyl)

# Print a sample of data
head(mtcars)
```

## ggplot Basics
### ggplot basic example
```{r}
# Basic scatter plot
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# Change the font size and shape
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
   geom_point(size = 1.5, shape = 18)
```

### aes_string()
aes_string() generates aesthetics from a string. This is particularly useful when writing functions that create plots cause strings can be used to define aesthetic mappings. Otherwise it would be needed to use a substitute to generate a call to aes().

```{r}
ggplot(data = mtcars, aes_string(x = "wt", y = "mpg")) +
   geom_point(size = 1.5, shape = 18)

ggpoints <- function(data, x_name, y_name)
{
  p <- ggplot(data = data, aes_string(x_name, y_name)) +
    geom_point(color = "red") +
    geom_smooth()
  
  return(p)
}

ggpoints(data = mtcars, x_name = "wt", y_name = "mpg")
```

### Visualise transformations of original dataset
Some plots visualise a transformation on the original data set. In this case, an alternative way to build a layer is to use stat_*() functions.

```{r}
set.seed(1234)
wdata = data.frame(
  sex = factor(rep(c("F", "M"), each = 200)),
  weight = c(rnorm(200, 55), rnorm(200, 58)))

# Use geometry function
ggplot(wdata, aes(x = weight)) + geom_density()

ggplot(wdata, aes(x = weight)) + stat_density()
```

### Layers using geoms
Each plot consists of one or more layers. Layers are defined using geom_*().

#### Using the same data and same aesthetic mapping in one plot
```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point() +  # to draw points
  geom_line()     # to draw a line
```

#### Using different data and mappings for different layers
```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point() +  # to draw points from whole data set
  geom_line(data = head(mtcars), color = "red")  # to draw a line from a subset of data
```

### Calculations in aes()
```{r}
ggplot(data = mtcars, aes(x = log2(wt), y = log2(mpg))) +
  geom_point()
```

### Saving graphic objects
#### Saving directly from screen
```{r}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
ggsave("myplot1.pdf")  # to PDF
ggsave("myplot1.png")  # to PNG
```

#### Saveing graphic objects
```{r}
pdf("myplot2.pdf")
myplot <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
print(myplot)
dev.off()

png("myplot2.png")
myplot <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
print(myplot)
dev.off()
```

