---
title: "Beautiful Graphics"
subtitle: "Notes to the book by Alboukadel Kassambara"
author:
- name: "Georg Maubach"
  affiliation: "Veikko Consulting"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    self_contained: yes
    toc: no
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    toc_title: "Table of Contents"
  pdf_document:
    keep_md: yes
    number_sections: yes
    toc: no
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_title: "Table of Contents"
  word_document:
    toc: no
    toc_depth: 3
keywords: graphics, notes, ggplot2
linestretch: 1
fontsize: 11pt
tags:
- graphics
- notes
- ggplot2
abstract: |
  This document contains notes while reading the book.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The [books web site](http://www.sthda.com/english/wiki/ggplot2-essentials).

# Introduction
## Terms
**Please note:**  
A plot can be divided into **3** fundumental parts:
**Plot = data + Aesthetics + Geometry**.

- **data**: a data frame
- **Aesthetics**: describe the mapping of variables to visual properties of geometry.
- **Geometry**: describe the type of graphic, e.g. histogram, box plot, line plot, scatter plot, etc.

### Further reading
- [Aesthetics Documentation](http://docs.ggplot2.org/0.9.3/aes.html)
- [Aesthetics Specification](http://docs.ggplot2.org/current/vignettes/ggplot2-specs.html)

## Installing and Loading ggplot2
```{r eval = FALSE}
warning("Use type = 'source' only if you have configured a C++ compiler.")

local({
  r = getOption("repos")
  r["CRAN"] = "https://cran.rstudio.com/"
  r["CRANextra"] <- "http://www.stats.ox.ac.uk/pub/RWin"
  r["CRAN_de"] <- "https://cran.uni-muenster.de/"
  options(repos = r)
})

install.packages(
  pkgs = "ggplot2", 
  dependencies = TRUE,
  type = "source")

library(ggplot2)
```

## Working with ggplot2
- **Please note:**  
  In order for ggplot2 to work properly the data should be tidy, i.e. observations shall be rows, columns should be the characteristics.
- The output of a plot can be kept as a variable and printed any time using the print() function.
- last_plot() returns the last created plot for modification.
- ggsave("plot.png", width = 5, height = 5) save the last plot in the current working directory.
